---
title: "Getting started with retirer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with retirer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 6,
  out.width = "100%",
  fig.align = "center",
  dpi = 300
)
```

# Introduction

`retirer` is designed for exploratory analysis of retirement-age and demographic data in a single package.

It combines effective labour market exit age data from the OECD with demographic indicators from the United Nations World Population Prospects. The package is especially useful for graduate-level exploratory work where you want to move back and forth between:

* time trends,
* one-year comparisons,
* women-versus-men analysis,
* relationships between measures, and
* simple geographic comparisons of OECD/EU membership groups.

Understanding trends of retirement age has become increasingly important among the dramatic aging of certain countries as well as overall increased life expectancy. This data can help clarify how countries are handling the prevalence of these trends and remedy potential issues such as shrinking working-age population and pressure on pensions. 

This vignette walks through a typical workflow.
You can find the source code here: [retirer GitHub repository](https://github.com/coopernelson112/retirer)

# Data provenance

The package data included in `retirer` are a cleaned and joined version of two public data sources:

* OECD effective labour market exit age data by country, sex, and year
* United Nations World Population Prospects 2024 demographic indicators

The package includes the cleaned analysis-ready data set, `retirement`, for live examples and reproducible analysis. The data-construction logic used during development is kept in `data-raw/build_retirement.R`.

The package is designed so that most users can work directly with the cleaned package data, while still being able to inspect how the final data set was assembled.

# The data

The package ships with a built-in data set called `retirement`.

```{r}
library(retirer)
dim(retirement)
names(retirement)
```

Each row represents one country or aggregate series in one year. The data include both country observations and aggregate entries such as `"European Union (27 countries)"` and `"OECD - Average"`.

You can inspect the available countries, groups, and measures with helper functions:

```{r}
retirement_groups()
retirement_countries()
retirement_variables()
sex_comparable_measures()
```

## Country names and aliases

The package accepts several common country-name aliases for convenience. For example, `"Turkey"` resolves to `"Türkiye"` and `"South Korea"` resolves to `"Korea"` in returned plot labels or data labels.

```{r}
retirement_country_aliases()
```

## Subsetting

This data contains an extensive amount of variables, of which not all may be useful towards your own research. The following function creates a secondary, smaller data set containing only the variables of your choosing. You may either select variables you would like to keep or exclude, and you may choose which data set to subset, though the default data set is `retirement`.

```{r}
choose_var()
```

# 1. Start with trends

A good first step is usually to inspect broad time trends.

## Retirement-age trends

The default retirement plot compares the four OECD/EU membership groups over time.

```{r, fig.cap = "Average Retirement Age trends for various EU and OECD Affiliations"}
plot_retirement_trends()
```
The plot shows that OECD country's average retirement age has been rising in recent years. Non affiliated countries have had a consistent downtrend over the last 50 years.

You can also focus on selected countries.

```{r, fig.cap = "Average retirement age trend for Japan, Turkey, and South Korea"}
plot_retirement_trends(
  level = "country",
  countries = c("Japan", "Germany", "United States")
)
```
All three countries have similar trend lines, with Japan having the highest average retirement age over the last 50 years.

## Demographic trends

Use `plot_demography_trends()` for demographic measures such as fertility, migration, mortality, life expectancy, or population size.

```{r, fig.cap = "Fertility rate trends for various EU and OECD Affiliations"}
plot_demography_trends(measure = "TFR")
```
There has been a sharp downturn in fertility rates among all country affiliations.

Life-expectancy measures also support sex-specific plotting.

```{r, fig.cap = "Life expextancy for 65 year-old women of Japan, Turkey, and South Korea"}
plot_demography_trends(
  measure = "LE65",
  sex = "women",
  level = "country",
  countries = c("Japan", "Turkey", "South Korea")
)
```
Life expectancy for both Japan and South Korea elderly women have sharply increased over the last 50 years, while Turkey women have modestly increased.

## Paired retirement-demography trends

If you want to compare retirement age with one demographic measure across time, use:

```{r, "Comparison between life expectancy of 65 year olds and retirement age for various country affiliations"}
plot_retirement_vs_demography(
  demographic_measure = "LE65"
)
```
Life-expectancy has steadily increased for all affiliations, while retirement age has returned to the same level it was 50 years ago approximately.

## Map-based group comparisons

In addition to time-series and cross-sectional plots, `retirer` can also display OECD/EU membership groups on a world map.

This is useful when your main question is geographic rather than temporal.

```{r, fig.cap= "World map of various country affiliations"}
map_groups()
```
Most affiliations are geographically based. 

You can also focus on selected groups and a specific year.

```{r, fig.cap = "World map of membership affiliations in the year 2020"}
map_groups(
  year = 2020,
  groups = c("OECD + EU", "OECD only"),
  include_nonrepresented = TRUE
)
```

# 2. Move to cross-sectional snapshots

Trend plots help you see change over time. Snapshot plots help you compare groups or countries in a single year.

## Snapshot data

`measure_snapshot_data()` returns the underlying one-year data frame.

```{r}
measure_snapshot_data(
  measure = "retirement_age",
  year = 2020,
  sex = "women"
)
```


## Snapshot plots

The corresponding plot is:

```{r, fig.cap = "Average retirement age for women in 2020 among country affiliations"}
plot_measure_snapshot(
  measure = "retirement_age",
  year = 2020,
  sex = "women"
)
```
Women in EU countries, which are not OECD members had the lowest average retirement age 
in 2020. 

Country-level snapshots are often useful for side-by-side comparison.

```{r, fig.cap = "Life expectancy for 65 year olds from China, Japan, Germany, and Turkey in 2020"}
plot_measure_snapshot(
  measure = "LE65",
  year = 2020,
  level = "country",
  countries = c("China", "Japan", "Germany", "Turkey"),
  sort = "descending"
)
```
Japanese elderly had a 7 year longer life expectancy than the lowest life expectancy, Turkey.

# 3. Analyze women-versus-men differences

A major use case for `retirer` is explicit sex comparison.

## Which measures support sex comparison?

Not every demographic variable has a women/men version. Use:

```{r}
sex_comparable_measures()
```

For sex-comparable demographic measures, `sex = "both"` uses the official combined-series column in the dataset. Retirement age is the main exception: because the data only provide separate women’s and men’s retirement ages, `sex = "both"` uses their arithmetic mean.

## Women and men over time

Use `plot_sex_comparison_trends()` when you want to compare the two series directly.

```{r, fig.cap = "Retirement age trends of men vs women among various country affiliations"}
plot_sex_comparison_trends(
  measure = "retirement_age"
)
```
Traditionally, men have a higher average retirement age than women.

You can also do this for other sex-comparable measures:

```{r, fig.cap = "Mortality before 40 for men and women in China, Japan, Germany, and Turkey"}
plot_sex_comparison_trends(
  measure = "Q0040",
  level = "country",
  countries = c("China", "Japan", "Germany", "Turkey")
)
```
Mortality for men and women are similar, with Turkey having a large decrease in mortality before age 40 in the last 50 years.
## Sex gaps over time

Use `plot_sex_gap_trends()` when you want a directed difference instead of two separate lines.

```{r, fig.cap = "Trend of retirement age gap among China, Japan, and Germany"}
plot_sex_gap_trends(
  measure = "retirement_age",
  level = "country",
  countries = c("China", "Japan", "Germany")
)
```
China has had the largest average retirement age gap between men and women over the last 50 years. 

You can reverse the sign convention if needed.

```{r}
plot_sex_gap_trends(
  measure = "LE65",
  gap = "men_minus_women"
)
```

Gap plots always stay on the original scale of the measure. For example, a gap in `retirement_age` is measured in years, while a gap in `Q1560` is measured on the original mortality scale used by that variable.

## Sex-gap snapshots

You can also compare sex gaps across groups or countries in a single year.

```{r}
sex_gap_data(
  measure = "LE65",
  year = 2020,
  level = "country",
  countries = c("China", "Japan")
)
```

```{r, fig.cap = "Difference in Age 15-60 mortalities between men and women in 2020"}
plot_sex_gap_snapshot(
  measure = "Q1560",
  year = 2020,
  level = "country",
  countries = c("China", "Japan", "Germany"),
  gap = "women_minus_men"
)
```
Men mortalities were greater then women mortalities for all countries, with China having the largest difference.

## Retirement gaps versus demographic gaps

If you want to compare the sex gap in retirement age to the sex gap in another measure, use:

```{r, fig.cap = "Life expectancy gender gap vs Retirement age gender gap among vaious country affiliations"}
plot_retirement_vs_demography_gap(
  demographic_measure = "LE65"
)
```
Women have higher life expectancy at age 65, with lower average retirement age.

# 4. Study relationships between measures

Once you understand trends and snapshots, the next step is often to ask whether two measures are associated.

## Relationship data

`measure_relationship_data()` returns a tidy one-year cross section with both measures.

```{r}
measure_relationship_data(
  x_measure = "LE65",
  y_measure = "retirement_age",
  year = 2020,
  level = "country",
  countries = c("China", "Japan", "Germany")
)
```

## Relationship plots

Use `plot_measure_relationship()` for a scatterplot.

```{r, fig.cap= "Scatterplot of Life Expectancy vs Retirement Age in 2020"}
plot_measure_relationship(
  x_measure = "LE65",
  y_measure = "retirement_age",
  year = 2020,
  level = "country",
  countries = c("China", "Japan", "Germany"),
  show_labels = TRUE,
  add_smooth = FALSE
)
```
Life expectancy seems relatively correlated with retirement age, Japan had by far the highest life expectancy and retirement age among countries listed.

You can also compare measures with different sex selections.

```{r, fig.cap= "Life Expectancy for Women vs Retirement Age for Men in 2020"}
plot_measure_relationship(
  x_measure = "LE65",
  y_measure = "retirement_age",
  year = 2020,
  x_sex = "women",
  y_sex = "men",
  level = "country",
  countries = c("China", "Japan", "Germany")
)
```
Higher life expectancy for women seems to be loosely correlated with higher retirement age for men in listed countries.

By default, `plot_measure_relationship()` adds a black least-squares regression line. Set `add_smooth = FALSE` when you want only the points.

## Gap relationships

Use `plot_gap_relationship()` when both axes are directed sex gaps.

```{r, , fig.cap = "Scatterplot of Life Expectancy Gender Gap vs Retirement Age Gender Gap in 2020"}
plot_gap_relationship(
  x_measure = "LE65",
  y_measure = "retirement_age",
  year = 2020,
  level = "country",
  countries = c("China", "Japan", "Germany"),
  gap = "women_minus_men",
  show_labels = TRUE,
  add_smooth = FALSE
)
```

This is useful when your substantive question is about disparities rather than levels.

# 5. A typical exploratory workflow

For a graduate-level exploratory project, a good sequence is:

1. Start with a trend plot to understand the broad time pattern.
2. Use a map when a geographic comparison is the most natural first view.
3. Use a snapshot plot to compare countries or groups in a focal year.
4. Check whether the measure supports sex comparison.
5. Inspect women-versus-men differences directly with comparison and gap plots.
6. Use relationship plots to see whether two measures move together.
7. Pull the underlying data with `*_data()` helpers when you need custom figures or tables.

For example:

```{r}
plot_retirement_trends(
  level = "country",
  countries = c("China", "Japan", "Germany")
)

plot_measure_snapshot(
  measure = "LE65",
  year = 2020,
  level = "country",
  countries = c("China", "Japan", "Germany")
)

plot_sex_gap_trends(
  measure = "retirement_age",
  level = "country",
  countries = c("China", "Japan", "Germany")
)

plot_measure_relationship(
  x_measure = "LE65",
  y_measure = "retirement_age",
  year = 2020,
  level = "country",
  countries = c("China", "Japan", "Germany"),
  show_labels = TRUE,
  add_smooth = FALSE
)
```

# 6. Custom analysis with the returned data

One of the strengths of `retirer` is that the plotting functions are paired with tidy data helpers.

For example, you can get snapshot data and then make your own plot:

```{r}
snap <- measure_snapshot_data(
  measure = "retirement_age",
  year = 2020,
  sex = "women",
  level = "country",
  countries = c("China", "Japan", "Germany")
)

snap
```

Or you can get relationship data and fit your own model:

```{r}
rel <- measure_relationship_data(
  x_measure = "LE65",
  y_measure = "retirement_age",
  year = 2020,
  level = "country",
  countries = c("China", "Japan", "Germany")
)

rel
```

# Conclusion

`retirer` is designed to support a coherent exploratory workflow rather than just isolated plotting calls.

The core idea is simple:

* use trend plots for change over time,
* use maps for geographic group comparison,
* use snapshot plots for one-year comparisons,
* use sex-comparison functions for women-versus-men analysis,
* use relationship plots to study associations, and
* use the data helpers whenever you need custom outputs.

That combination should cover many first-pass research tasks in retirement and demographic analysis.
